I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-26 Paper/proofs.tex
th3
th2

\begin{proof}
	We show that $\beta=\max(z)$.

@@

\subsection{From weak to compensating}

			We now show Theorem \ref{th2} from Theorem \ref{thm:2}.
			Assume that $n$ is a node in a weakly compensating path $\rho$.

			There are three mutually exclusive cases:
			\begin{enumerate}
				\item $n$ is on a compensating pair of paths $\rho,\rho'$ to $z$ (but not its source or target), otherwise:

				\item $n$ is on a compensating pair of paths $\rho,\rho'$ to $z' \neq z$ (but not its source or target), otherwise:

				\item $n$ is not on any compensating pair of paths to any node (except may be as source or target).
			\end{enumerate}

@@

			In the other cases, the ReLU associated with node $n$ will not be in $X$ 
			(not opened in $\MILP_X$).
			In the second case, as we assumed without loss of generality that $z$ was the unique output node, it means that $z'$ is in a layer strictly before $z'$.
			Consider all the possible associated $z'$.

			In the last case, $a$ is in a compensating path (the difference is that this path is not from the input to output), and will be opened. So we focus on the first and second case.

			For the second case, if we replace $P'$ with $P''=P'_0,\cdots,P'_{i-1},P_{i},P'_{i+1},\cdots$, then $P,P''$ is a weak compensating pair that $a$ is a common nodes. So we only need to consider case 1.

@@

			Combining all discussion above, if $a$ is in a weak compensating path from input to output, then it is in a compensating path from one layer to another layer. Therefore, Theorem \ref{thm:2} implies Theorem \ref{th2}. This is what we want to show.

>>>>

\begin{proof}[of Theorem \ref{thm:2}]
	We show that $\beta=\max(z)$.

@@

\subsection{Proof of Theorem \ref{th2}}

Note: it does not need directly the previous proof, although it uses the same arguments.
We should adapt the previous argument here, and then remove B.1. altogether.

We assume the fairly light hypothesis (H1) that the DNN is such that for every 2 paths $\rho_1,\rho_2$ (not necessarily disjoint) from any $m$ to any $n$ at least 2 layers apart, there exists another path $\rho_3$ from $m$ to $n$ (with non zero-weight) totally disjoint from $\rho_1$ and $\rho_2$. Further we assume without loss of generality that every node can be reached from at least one input with a path of non-zero weight (otherwise we just remove this node altogether).
We also assume without loss of generality that there is no node which can only reach $z$ with path of weight $0$, as such nodes have no contribution to the value $\val(z)$. We denote $m<n$ to mean that there exists a path with non zero weight from $m$ to $n$.

			We now show Theorem \ref{th2} from Theorem \ref{thm:2}.
			Let $X$ be the set of nodes on compensating paths with target $z$, but not as source or target.

			Let $Z$ be the subset of nodes $n$ of the DNN such that $S$ is defined on $n$ 
			wrt node $z$. As proved earlier, $Z$ is suffix closed, i.e. if $m<n<z$ with $m \in Z$, then $n \in Z$. We can partition $Z = Z_{+} \sqcup Z_{-}$, for positive and negative sign. They form 2 independent subnetworks of the DNN.
			$X$ can intersect both $Z_+$ and $Z_-$.
			For nodes not in $Z$: some may be in $X$, and some may be out of $X$.
			We call $X_+ = Z_+ \cap X$ and $X_-=Z_- \cap X$, as well as 
			defined $Y_+,Y_-$ such that $Z_+ = X_+ \sqcup Y_+$ and $Z_- = X_- \sqcup Y_-$.

			Because of (H1), nodes not in $Z$ (and not in the inputs) are in $X$:
			node $n \notin Z$ implies the existence of two path $\rho_1,\rho_2$ (not necessarily disjoint) to $z$, with opposite sign. 
			Take an input node $a$ reaching $n$ with path $\rho$ with non zero weight.
			Then using (H1), we get a third path $\rho_3$ disjoint from $\rho \cdot \rho_1$ and from $\rho \cdot \rho_2$ with non-zero weight. Depending on the sign of $weight(\rho_3)$, it will form a compensating pair of path to $z$ either with $\rho \cdot \rho_1$ or with $\rho \cdot \rho_2$, hence $n$ will be in $X$. We call $X_*$ this subset $X$.

\begin{lemma}
	\label{lemma10}
		Let $y \in Y_+$. Then for all $m$, there does not exist a pair $(\rho_1,\rho_2)$ 
		of paths  with weights of opposite sign with $\rho_1$ from $m$ to $y$ and 
		with $\rho_2$ from $m$ to $z$ (even non disjoint).
\end{lemma}

\begin{proof}
  $y \notin X$, so it is not on a pair of compensating pair with target $z$.
  Let $\rho$ be a path from $y$ to $z$ (it has positive weight has $y \in Z_+$).
  Hence $\rho_1 \rho, \rho_2$ have opposite sign, and they must meet before $z$.

  If $\rho_1 \rho$ meets $\rho_2$ after $y$, then we use $(H_1)$ to get another disjoint path from $y$ to $z$ (it has positive weight has $y \in Z_+$), it would not meet $\rho_2$, a contradiction.

  Hence $\rho_1$ must meet $\rho_2$ (that is before or at $y$).
  Their weights must be of opposite sign, otherwise we can inductively reapplied the same argument to their suffixes.
  Let $\rho_1=\rho_1'\rho_1''$.
  Both $\rho_1 \cdot \rho$ and $\rho_2' \cdot \rho_1'' \cdot \rho$
  are from $m$ to $z$ going through $y$, but with weights of opposite sign (as $\rho'_1$ and $\rho'_2$ have opposite sign).
  Now, by $(H1)$, there exists a path $\rho_3$ disjoint from $\rho_1 \cdot \rho$ and$\rho_2' \cdot \rho_1'' \cdot \rho$ from $m$ to $z$. It means that we have a pair of compensating paths from $m$ tp $z$, and thus $y \in X$, a contradiction with $y \in Y_+$. \qed
\end{proof}

It means that if a path $\rho_1$ from $m$ to $y$ has stricly positive (resp. negative) weight, then all the paths from $m$ to $z$ will also have strictly positive (resp. negative) weights. It means that Sign is defined on all the $m$ that can reach $Y_+$.

We have the same results for $y \in Y_-$, for paths $\rho_1, \rho_2$ of same sign:

\begin{lemma}
	\label{lemma11}
	Let $y \in Y_-$. Then for all $m$, there does not exist a pair $(\rho_1,\rho_2)$ 
	of paths with weights of same sign with $\rho_1$ from $m$ to $y$ and 
	with $\rho_2$ from $m$ to $z$ (even non disjoint).
\end{lemma}

Hence, we obtain as corollary:

\begin{corollary}
\label{cor:simple}
Assume that $m < n$ with $n \in Y= Y_+ \sqcup Y_-$.
Then $m \in Z$.
\end{corollary}

From there, we conclude that $Y$ is actually prefix-closed:

\begin{lemma}
	\label{lemma12}
	Assume that $m < n$ with $n \in Y$.
	Then $m \in Y$.
\end{lemma}

\begin{proof}
Assume by contradiction that $m \notin Y$ with $m<n \in Y$.
That is, $m \in X$, and there exists a node $m'$ with a pair of compensating paths to $z$
passing by $m$. But now, that means that $m'<m<n<z$.
Thus $m'$ is in $Z$, a contradiction with $m'$ the source of a pair of compensating paths.
\qed
\end{proof}

As in the proof of Theorem \ref{thm:2}, we can then optimize $Y$ first on its own (considering the sign of every node), and consider its (direct and indirect) contribution to $z$ as biases, before optimizing the rest of the network.
This is exactly what $\MILP_X$ is doing:
the bias from each node of $Y$ comes from the bounds on nodes (of $Y$) computed inductively in the previous layers, and then there is a global optimization of the rest of the network (set of nodes $X$ whose ReLU are encoded exactly in $\MILP_X$).

This concludes the proof of Theorem \ref{th2}

\iffalse

\begin{lemma}
	There does not exist a node $m$ with $m<n^+ \in Y^+$ and $m<n^- \in Y^-$.
	\end{lemma}

	\begin{proof}
	By contradiction, assume that there exist nodes $m, n^+ \in Y_+$ and  $n^
	- \in Y_-$, such that there are two non-zero weight paths $\rho_1$ from $m$ to $n^+$ and $\rho_2$ from $n$ to $n^-$ Assume first that the weights of $\rho_1,\rho_2$ have the same sign, 
	in which case it gives us a path $\rho_2 \rho_3$  from $m$ to $z$ of opposite sign than $\rho_1$. Applying Lemma \ref{lemma10}, we have prefixes $\rho'_1,\rho'_2$ of $\rho_1,\rho_2$ ending in the same state $m'$, and with weights of opposite sign.
	So we are left with path from $m'$ to $Y_-,Y_+$ of opposite sign. 
	We then apply Lemma \label{lemma11}, inductively then applying Lemma \label{lemma10} on the suffices etc. We obtain a contradiction to the existence of $m$ with $m<n^+$ and $m<n^-$. \qed
	\end{proof}

	With the same arguments, we obtain:

	\begin{lemma}
	If we have $m<n \in X_*$ and $m < n^+ \in Y_+$ (or $m < n^- \in Y_-$),
	then $n < n^+$ (or $n < n^-$), with compensating pairs of path from $m$ to $n$.
	\end{lemma}

	Last, consider the case $m<n^+ \in Y_+$ and $m < n^- \in X_-$.

@@

			In the other cases, the ReLU associated with node $n$ will not be in $X$ 
			(not opened in $\MILP_X$).
			In the second case, as 
			, it means that $z'$ is in a layer strictly before $z'$.
			Consider all the possible associated target nodes $z'$ of compensating path from $n$.

@@

			Combining all discussion above, if $a$ is in a weak compensating path from input to output, then it is in a compensating path from one layer to another layer. Therefore, Theorem \ref{thm:2} implies Theorem \ref{th2}. This is what we want to show.
			\fi
